---
title: "Reproducible Research Course Project 1"
author: "SMK"
date: "5/10/2022"
output: html_document
---

<p>&nbsp;</p>
<p>&nbsp;</p>

**This R Markdown file is for the Course Project 1 of Reproducible Research course from Coursera.**

**The purpose of this file is to use the dataset from file 'activity.csv' and process the data and answer the questions related to this dataset such as loading the data, creating graphs and imputing missing data.**


<p>&nbsp;</p>

***Set global option to echo R code***

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<p>&nbsp;</p>

#### 1. Code for reading in the dataset and/or processing the data

***Read input .csv file and display summary***

```{r}
unzip("./activity.zip")
activity_data <- read.csv("./activity.csv")
head(activity_data)
```

```{r}
summary(activity_data)
```

***Calculate total number of steps taken per day***

```{r}
no_of_steps <- aggregate(steps~date, activity_data, sum, na.rm=TRUE)
head(no_of_steps)
```

<p>&nbsp;</p>

#### 2. Histogram of the total number of steps taken each day

```{r}
png("Q_2_Histogram_Total_Steps.png")
hist(no_of_steps$steps, xlab="Number of steps per day", ylab="Frequency", main="Total number of steps taken each day", col="light blue")
dev.off()
```

<p>&nbsp;</p>

#### 3. Mean and median number of steps taken each day

* Mean

```{r}
mean_steps <- mean(no_of_steps$steps)
print(paste("Mean number of steps taken each day:", round(mean_steps, digits=2)))
```

* Median

```{r}
median_steps <- median(no_of_steps$steps)
print(paste("Median number of steps taken each day:", median_steps))
```

<p>&nbsp;</p>

#### 4. Time series plot of the average number of steps taken

```{r}
steps_interval <- aggregate(steps~interval, data=activity_data, mean, na.rm=TRUE)
png("Q_4_Time_Series_Average_Steps.png")
plot(steps~interval, data=steps_interval, type="l", col="brown")
dev.off()
```

<p>&nbsp;</p>

#### 5. The 5-minute interval that, on average, contains the maximum number of steps

```{r}
max_steps_interval <- steps_interval[which.max(steps_interval$steps),]$interval
print(paste("The 5-minute interval that contains the maximum number of average steps:", max_steps_interval))
```

<p>&nbsp;</p>

#### 6. Code to describe and show a strategy for imputing missing data

***Number of missing values***

```{r}
no_of_missing_val <- sum(is.na(activity_data$steps))
print(paste("The 5-minute interval that contains the maximum number of average steps:", no_of_missing_val))
```

***Filling in all of the missing values in the dataset with 'average'***

```{r}
missing_steps_val <- mean(steps_interval$steps)
missing_steps_val
```

***New dataset with missing data***

```{r}
activity_fill_missing <- activity_data
missing_val_idx <- is.na(activity_fill_missing[,1])
activity_fill_missing[missing_val_idx, 1] <- missing_steps_val
head(activity_fill_missing)
```

<p>&nbsp;</p>

#### 7. Histogram of the total number of steps taken each day after missing values are imputed

```{r}
no_of_steps_new <- aggregate(steps~date, activity_fill_missing, sum, na.rm=TRUE)
png("Q_7_Histogram_Total_Steps_After_Imputed_Missing_Values.png")
hist(no_of_steps_new$steps, xlab="Number of steps per day", ylab="Frequency", main="Total number of steps taken each day after imputing missing values", col="light blue")
dev.off()
```

* Mean of the total number of steps taken per day with new dataset after missing values are imputed

```{r}
mean_steps_new <- mean(no_of_steps_new$steps)
print(paste("Mean number of steps taken each day after imputing missing values:", round(mean_steps_new, digits=2)))
```

* Median of the total number of steps taken per day with new dataset after missing values are imputed

```{r}
median_steps_new <- median(no_of_steps_new$steps)
print(paste("Median number of steps taken each day after imputing missing values:", round(median_steps_new, digits=2)))
```

***Mean of the data set after filling in missing values remains the same as the missing values are filled in using average values. Median has changed little bit after imputing the missing values***

<p>&nbsp;</p>

#### 8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

***New factor variable in the dataset with two levels – "Weekday" and "Weekend"***

```{r}
activity_fill_missing$date <- as.Date(strptime(activity_fill_missing$date, format="%Y-%m-%d"))
activity_fill_missing$day <- weekdays(activity_fill_missing$date) 
for (i in 1:nrow(activity_fill_missing))
{
  if(activity_fill_missing[i,]$day %in% c("Saturday", "Sunday"))
  {
    activity_fill_missing[i,]$day <- "Weekend"
  }
  else
  {
    activity_fill_missing[i,]$day <- "Weekday"
  }
}
head(activity_fill_missing)
```

***Time series plot of the average number of steps taken across weekday days and weekend days***

```{r}
library(ggplot2)
steps_new_by_day <- aggregate(steps~interval+day, activity_fill_missing, mean, na.rm=TRUE)
png("Q_8_Panel_Plot_Average_Steps_Weekdays_vs_Weekends.png")
plot <- ggplot(steps_new_by_day, aes(x=interval, y=steps, color=day)) +
          geom_line() +
          labs(title="Average Number of Steps By Day", x="Interval", y="Average Number of Steps") +
          facet_wrap(~day, ncol=1, nrow=2) +
          theme(plot.title = element_text(hjust = 0.5))
print(plot)
dev.off()
```

***The time series plots of Weekday vs Weekend shows that for Weekday maximum walking was done in the interval of 500-1000 and for Weekend maximum walking was done in the interval of 800-1700.***
